---
title: 'Personal Activity Efficacy: Personal Activity Device Data Analysis'
author: "Robert Green"
date: "01/05/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. This project will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). [^1]

[^1]: Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 
Cited by 2 (Google Scholar)

The data has been downloaded from the website and save as 'pml-training.csv' and 'pml-testin.csv', respectively.

## Prepare data for analysis

``` {r trainingDataLoad, message = FALSE}
library(caret); library(kernlab); library(RANN); library(randomForest); library(stringr)
train.data <- read.csv('pml-training.csv')
dim(train.data)
str(train.data, list.len=20)
```
We can see the training dataset consists of 160 variables across 19,622 observations. 

Looking specifically at our response variable 'classe', we see it is a factor.
``` {r factor_classe}
str(train.data$classe)
```

Let's also eliminate the timestamp and name columns, assuming that time-of-day did not influence it the excercises were done correctly, and any columns that are 'summary'-type (i.e., average) columns.
``` {r eliminate_TOD_vars}
col_list <- colnames(train.data)
rem_col_list <- col_list[ str_detect( col_list, "(var|stddev|avg|total)" ) ]
rem_col_list <- c(rem_col_list, "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "user_name", "X", "num_window")
train.data.noTOD <- train.data[ !col_list %in% rem_col_list]
```

There are also quite a few columns that are mostly blank, so let's eliminate those as well. Turns out they are factor columns, so I will eliminate all columns that are of type "factor" with the exception of classe.
```{r remove_factor_cols, results = 'hide'}
## dat[, -which(colMeans(is.na(dat)) > 0.5)]
use_col <- NULL
for (col_name in col_list)
{ if (is.numeric( train.data.noTOD [[ col_name ]] ) == T | is.integer ( train.data.noTOD [[ col_name ]] ) == T )
use_col <- append(col_name, use_col)}
use_col2 <- c(use_col, "classe")
train.data.tidy <- train.data.noTOD[ , use_col2]
```
Columns that have more than 50% NA values are not going to give much guidance. I will eliminate them as well:
``` {r rem_col_NA}
train.data.noNA <- train.data.tidy[, -which(colMeans(is.na(train.data.tidy)) > 0.5)]
```

Clean up unused data and run garbage collection:
``` {r clean_up_time}
rm(train.data)
rm(train.data.noTOD)
gc()
```

## Analysis
To apply cross-validation, I will use k-folds. Defining a control:
``` {r control}
set.seed(32323)
control <- trainControl(method = "cv", number = 5)
```

Start with a random forest model to train the prediction model using the k-folds control:
``` {r rf_mod, results = 'hide'}
rf_mod <- train(classe ~ ., data = train.data.noNA, method="rf", trControl = control)
gc()
``` 

Build a gbm model for comparison:
``` {r gbm_mod_1, message = FALSE, results = 'hide'}
gbm_mod <- train(classe ~ ., data = train.data.noNA, method="gbm", trControl = control)
gc()
``` 

## Data Analysis Conclusions
Looking at the two models:
``` {r compare}
rf_mod
gbm_mod
```

I would use the random forest method for the testing data to predict the results.  This is based on the better predicted performance of the random forest model than the gbm model.

## Application
I now need to pull in the test data and perform the same data clean up operations, and apply the model to the test data to predict the manner in which the subjects did the exercise. For sake of brevity, the r code is shown in one block, and is essentially the same clean up done for the training data.
```{r test_data_cleanup}
test.data <- read.csv('pml-testing.csv')
test.data.noTOD <- test.data[ !col_list %in% rem_col_list]
test.data.tidy <- test.data.noTOD[ , use_col]
test.data.noNA <- test.data.tidy[, -which(colMeans(is.na(train.data.tidy)) > 0.5)]
```

Applying the rf_mod to the testing data:
```{r pred}
predictions <- predict(rf_mod, test.data.noNA)
answers <- data.frame(predictions)
answers
```
